Monte Carlo Synthetic Acceleration Solutions for the Simplified $P_N$ Approximation

### 2014-01-16 doc/ttsp2013/ttsp_2013.tex
updating conclusion to ttsp paper

Predictive modeling and simulation capability requires the combination
of high fidelity models, high performance computing hardware that can
handle the intense computational loads required by these models, and
modern algorithms for solving these problems that leverage this high
performance hardware. For nuclear reactor analysis, this predictive
capability can enable tighter design tolerances for improved thermal
performance and efficiency, higher fuel burn-up and therefore
reduction in generated waste, and high confidence in accident scenario
models. Modern high fidelity deterministic methods for large scale
problems are often variants on the discrete ordinates ($S_N$) method
\cite{evans_denovo:_2010}. For fission reactor neutronics simulations,
the $S_N$ method requires potentially trillions of unknown angular
flux moments to be computed to achieve good accuracy for the responses
of interest \cite{slaybaugh_acceleration_2011}.

@@

In previous work \cite{evans_monte_2012}, we applied the Monte Carlo
synthetic acceleration (MCSA) method the thermal radiation diffusion
equation. For several transient problems we observed that MCSA
performed competetively with preconditioned Krylov methods. In a
continuation of these studies, we have applied the method light water
reactor problems discretized by the $SP_N$ equations. This system has
no time dependence, a large scattering component, and assymetry in the
linear operator, creating a set of problems complementary to the
transient, opaque and symmetric systems of our previous analyis.

In order to leverage Monte Carlo Synthetic Acceleration as a solution
technique for neutron transport problems, its current formulation
requires the linear operator and all preconditioners to be explicitly
formed. Recent developments in the Exnihilo neutronics package at Oak
Ridge National Laboratory have permitted generation of the $SP_N$
system of equations for detailed neutronics models of fission
reactor-based systems\cite{evans_simplified_2013}. By fully forming
these equations and formulating them as a linear algebra problem
instead of using an explicit iterative method, we now have access to
modern advancements in computational linear algebra including Krylov
solvers for asymmetric systems and algebraic preconditioning methods.

@@

\subsection{MCSA Solution of the $SP_N$ Equations}
\label{subsec:mcsa_solution}
Using the ideas of Halton, Evans and Mosher recently developed a Monte
Carlo solution method that was not prohibited severely by the quality
of the initial guess for the system \cite{evans_monte_2009} and later
applied it more rigorously as a solution mechanism for the radiation
diffusion equation \cite{evans_monte_2012}. With their new methods,
they achieved identical numerical results as conventional Krylov
solvers as well as comparable performance in both number of iterations
and CPU time. Their approach was instead to use residual Monte Carlo
as a synthetic acceleration for a stationary method. The
\textit{Fixed-Point Monte Carlo Synthetic Acceleration} (MCSA) method
is defined as:
\begin{subequations}
  \begin{gather}
    \ve{x}^{k+1/2} = \ve{x}^k + \ve{r}^k\:,\\
    \ve{r}^{k+1/2} = \ve{b} - \ve{A}\ve{x}^{k+1/2}\:,\\
    \ve{A}\delta\ve{x}^{k+1/2} = \ve{r}^{k+1/2}\:,\\
    \ve{x}^{k+1} = \ve{x}^{k+1/2} + \delta \ve{x}^{k+1/2}\:,\\
    \ve{r}^{k+1} = \ve{b} - \ve{A}\ve{x}^{k+1}\:,
  \end{gather}
  \label{eq:mcsa}
\end{subequations}
where a Neumann-Ulam Monte Carlo method is used to generate the
solution correction from the residual and Richardson's iteration in
the first step has been rewritten as a residual correction. Using
Monte Carlo in this way achieves the same effect as Halton's method,
decoupling its convergence rate from the overall convergence rate of
the method. Here, the approximate Monte Carlo solution is not driven
to a particular convergence as it merely supplies a correction for the
initial guess generated by Richardson's iteration. Rather, only a set
number of histories are required using the Neumann-Ulam method to
generate the correction. We refer the reader to
\cite{evans_monte_2012} for a more rigorous defintion of the method.

To generate the multiplication factor and steady-state flux
distribution for this problem, at every eigenvalue iteration MCSA is
used to solve the resulting $SP_N$ problem using the provided fission
source. Algorithm~\ref{alg:power_iteration} presents the use of MCSA
within a power iteration strategy to find the multiplication factor.
\begin{algorithm}[h!]
  \caption{Power Iteration MCSA Scheme}
  \label{alg:power_iteration}
  \begin{algorithmic}
    \State $k_0 =$ initial guess
    \State $\mathbf{\Phi}_0 =$ initial guess
    \State $n = 0$
    \While{$|\frac{k^n - k^{n-1}}{k^n}| < \epsilon$}
    \Comment{Iterate until convergence of the eigenvalue}
    \State $\mathbf{M} \mathbf{\Phi}^{n+1} = \frac{1}{k^n} \mathbf{F} \mathbf{\Phi}^n$
    \Comment{Solve for the new flux state with MCSA}
    \State $k^{n+1} = k^n \frac{\int \mathbf{F} \mathbf{\Phi}^{n+1} d\mathbf{r}}{\int
      \mathbf{F} \mathbf{\Phi}^n d\mathbf{r}}$
    \Comment{Update the multiplication factor}
    \State $n = n+1$
    \EndWhile
  \end{algorithmic}
\end{algorithm}
Here, $\mathbf{M}$ is the transport operator generated on the
left-hand side of the $SP_N$ discretization, $\mathbf{F}$ is the
fission matrix, and $\mathbf{\Phi}$ the multigroup neutron flux. This
problem is significantly more complicated than the simple test problem
used for the previous spectral analysis. Fission has been introduced
into the set of equations and the addition of moderator into the
system will increase the amount of scattering, creating a
significantly more difficult problem manifesting itself in an
iteration matrix with a larger spectral radius. When using MCSA, the
linear operator applied to $\mathbf{\Phi}^{n+1}$ at each eigenvalue
iteration will dictate convergence and remain unchanged throughout the
computation while the addition of fission to the system will only
modify the source of neutrons and the multiplication factor while not
affecting Monte Carlo transport.

@@

@@

In this paper, the Monte Carlo Synthetic Acceleration method has been
applied to the $SP_N$ discretization of the neutron transport equation
and explored in the context of a difficult nuclear fuel assembly
criticality problem. We observed that MCSA can solve the asymmetric
system generated by the $SP_N$ equations by incorporating the solution
technique into the Exnihilo neutronics code base developed at Oak
Ridge National Laboratory. In contrast to the thermal radiation
diffusion problem addressed in our previous work, light water reactor
problems are difficult to solve with MCSA as they have large spectral
radii due to the neutron scattering in the moderator. Because of this,
simple Jacobi-based preconditioning reduces the spectral radius of the
$SP_N$ system below unity, however, this alone was discovered not to
be sufficient criteria alone for MCSA convergence. As the spectral
radius approaches unity, MCSA was demonstrated to break down with more
stochastic histories required for convergence and more CPU time
required per history. To alleviate these effects, a left-right
preconditioned form of the MCSA method was developed and applied to
the $SP_N$ equations to obtain convergence. For the nuclear fuel
assembly problem, MCSA was observed to converge in fewer iterations
per eigenvalue iteration than GMRES for the fuel assembly criticality
problem and more than Bi-CGStab using the same
preconditioning. However, the explicit preconditioning strategy
required to overcome the MCSA spectral radius restriction creates run
times $O(100)$ larger than the production Krylov methods for the fuel
assembly criticality problem, indicating that significant research in
the area of Monte Carlo preconditioning implementations is required in
order for the method to be competetive in terms of time to solution.

>>>>

Predictive modeling and simulation capability requires the combination
of high fidelity models, high performance computing hardware that can
handle the intense computational loads required by these models, and
modern algorithms for solving these problems that leverage this high
performance hardware. For nuclear reactor analysis, this predictive
capability can enable tighter design tolerances for improved thermal
performance and efficiency, higher fuel utilization, and high
confidence in accident scenario models. Modern high fidelity
deterministic methods for large scale problems are often variants on
the discrete ordinates ($S_N$) method \cite{evans_denovo:_2010}. For
fission reactor neutronics simulations, the $S_N$ method requires
potentially trillions of unknown angular flux moments to be computed
to achieve good accuracy for the responses of interest
\cite{slaybaugh_acceleration_2011}. In the 1960's, Gelbard developed
an ad-hoc, multidimensional extension of the simple single dimension
planar $P_N$ equations that created a system of coupled,
diffusion-like equations known as the simplified $P_N$ ($SP_N$)
equations (cite gelbard). The reduction in numerical complexity
compared to current deterministic solution methods using the $S_N$
approximation could mean significant savings in both compute time and
memory required.

@@

In previous work \cite{evans_monte_2012}, we applied the Monte Carlo
synthetic acceleration (MCSA) method the thermal radiation diffusion
equation. For several transient problems we observed that MCSA
performed competetively with preconditioned Krylov methods. In a
continuation of these studies, we have applied the method light water
reactor problems discretized by the $SP_N$ equations. This system has
no time dependence, a large scattering component, and assymetry in the
linear operator, creating a set of problems complementary to the
transient, opaque, and symmetric systems of our previous analyis. In
addition, solving the $SP_N$ equations in this way also breaks away
from the $S_N$ forms of parallelism where spatial parallelism is
achieved by an efficient parallel sweep, angular efficiency achieved
by pipe-lining, and energy parallelism achieved by decoupling the
groups. With the $SP_N$ equations as a full matrix system, we now can
parallelize the problem as prescribed by the Monte Carlo algorithm,
which may be significantly more scalable than current $S_N$ transport
practices.

@@

\subsection{MCSA Solution of the $SP_N$ Equations}
\label{subsec:mcsa_solution}
To generate the multiplication factor and steady-state flux
distribution for this problem, at every eigenvalue iteration MCSA is
used to solve the resulting $SP_N$ problem using the provided fission
source. Algorithm~\ref{alg:power_iteration} presents the use of MCSA
within a power iteration strategy to find the multiplication factor.
\begin{algorithm}[h!]
  \caption{Power Iteration MCSA Scheme}
  \label{alg:power_iteration}
  \begin{algorithmic}
    \State $k_0 =$ initial guess
    \State $\mathbf{\Phi}_0 =$ initial guess
    \State $n = 0$
    \While{$|\frac{k^n - k^{n-1}}{k^n}| < \epsilon$}
    \Comment{Iterate until convergence of the eigenvalue}
    \State $\mathbf{M} \mathbf{\Phi}^{n+1} = \frac{1}{k^n} \mathbf{F} \mathbf{\Phi}^n$
    \Comment{Solve for the new flux state with MCSA}
    \State $k^{n+1} = k^n \frac{\int \mathbf{F} \mathbf{\Phi}^{n+1} d\mathbf{r}}{\int
      \mathbf{F} \mathbf{\Phi}^n d\mathbf{r}}$
    \Comment{Update the multiplication factor}
    \State $n = n+1$
    \EndWhile
  \end{algorithmic}
\end{algorithm}
Here, $\mathbf{M}$ is the transport operator generated on the
left-hand side of the $SP_N$ discretization, $\mathbf{F}$ is the
fission matrix, and $\mathbf{\Phi}$ the multigroup neutron flux. This
problem is significantly more complicated than the simple test problem
used for the previous spectral analysis. Fission has been introduced
into the set of equations and the addition of moderator into the
system will increase the amount of scattering, creating a
significantly more difficult problem manifesting itself in an
iteration matrix with a larger spectral radius. When using MCSA, the
linear operator applied to $\mathbf{\Phi}^{n+1}$ at each eigenvalue
iteration will dictate convergence and remain unchanged throughout the
computation while the addition of fission to the system will only
modify the source of neutrons and the multiplication factor while not
affecting Monte Carlo transport.

At each eigenvalue iteration, we then solve the linear system with the
MCSA:
\begin{subequations}
  \begin{gather}
    \ve{r}_{l} = \frac{1}{k^n}\ve{F}\mathbf{\Phi}^n -
    \ve{M}\mathbf{\Phi}^{n+1}_{l}\:,\\
    \mathbf{\Phi}^{n+1}_{l+1/2} = \mathbf{\Phi}^{n+1}_l + \ve{r}_l\:,\\
    \ve{r}_{l+1/2} = \frac{1}{k^n}\ve{F}\mathbf{\Phi}^n -
    \ve{M}\mathbf{\Phi}^{n+1}_{l+1/2}\:,\\
    \ve{M}\delta\mathbf{\Phi}^{n+1}_{l+1/2} = \ve{r}_{l+1/2}\:,\\
    \mathbf{\Phi}^{n+1}_{l+1} = \mathbf{\Phi}^{n+1}_{l+1/2} + \delta
    \mathbf{\Phi}_{l+1/2}\:,
  \end{gather}
  \label{eq:mcsa}
\end{subequations}
where the $l$ subscript indicates the MCSA iteration number and $n$
again indicates the eigenvalue iteration. We refer the reader to
\cite{evans_monte_2012} for a more rigorous defintion of the method.

@@

\subsection{Alternative Fixed Point Iterations}
\label{subsubsec:alternative_fixed_point}
In addition to preconditioning the basic Richardson iteration, the
MCSA algorithm presented in Eq~(\ref{eq:mcsa}) can be used to
accelerate any fixed point iteration which only depends on the state
(typically the residual) of the last iteration. As an example, for
positive-definite but not necessarily symmetric problems, the minimal
residual iteration \cite{saad_iterative_2003} can be used in
conjunction with MCSA:
\begin{subequations}
  \begin{gather}
    \alpha = \frac{\langle \mathbf{A}\mathbf{r}^k, \mathbf{r}^k
      \rangle}{\langle \mathbf{A}\mathbf{r}^k, \mathbf{A}\mathbf{r}^k
      \rangle}\:, \\
    \ve{x}^{k+1/2} = \ve{x}^k + \alpha \ve{r}^k\:,
    \label{eq:min_res_correction}\\
    \ve{r}^{k+1/2} = \ve{b} - \ve{A}\ve{x}^{k+1/2}\:,\\ 
    \ve{A}\delta\ve{x}^{k+1/2} = \ve{r}^{k+1/2}\:,\\ 
    \ve{x}^{k+1} = \ve{x}^{k+1/2} + \delta\ve{x}^{k+1/2}\:,\\ 
    \ve{r}^{k+1} = \ve{b} - \ve{A}\ve{x}^{k+1}\:.
  \end{gather}
  \label{eq:mcsa_min_res}
\end{subequations}
where $\alpha$ is an optimal extrapolation parameter generated from
the constraints. Interestingly, this scheme is nearly identical to the
Richardson iteration version in Eq~(\ref{eq:mcsa}) except now the
additional extrapolation parameter, $\alpha$, is applied to the
residual correction in Eq~(\ref{eq:min_res_correction}) as a means of
selecting a more optimal search direction at each iteration,
potentially further improving convergence.

@@

In this paper, the Monte Carlo Synthetic Acceleration method has been
applied to the $SP_N$ discretization of the neutron transport equation
and explored in the context of a difficult nuclear fuel assembly
criticality problem. We observed that MCSA can in fact solve the
asymmetric system generated by the $SP_N$ equations. In contrast to
the thermal radiation diffusion problem addressed in our previous
work, light water reactor problems are difficult to solve with MCSA as
they have large spectral radii due to the neutron scattering in the
moderator. Because of this, simple Jacobi-based preconditioning
reduces the spectral radius of the $SP_N$ system below unity, however,
this alone was discovered not to be sufficient criteria alone for MCSA
convergence. As the spectral radius approaches unity, MCSA was
demonstrated to break down with more stochastic histories required for
convergence and more CPU time required per history. To alleviate these
effects, a left-right preconditioned form of the MCSA method was
developed and applied to the $SP_N$ equations to obtain
convergence. For the nuclear fuel assembly problem, MCSA was observed
to converge in fewer iterations per eigenvalue iteration than GMRES
for the fuel assembly criticality problem and more than Bi-CGStab
using the same preconditioning. However, the explicit preconditioning
strategy required to overcome the MCSA spectral radius restriction
creates run times $O(100)$ larger than the production Krylov methods
for the fuel assembly criticality problem, indicating that significant
research in the area of Monte Carlo preconditioning implementations is
required in order for the method to be competetive in terms of time to
solution.
